\texttt{MSnbase}, efficient and elegant R-based processing and visualisation of raw mass spectrometry data

### 2020-07-15 msnbase2.Rnw
addressing more reviewer comments

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. We focus on the new \textit{on-disk} infrastructure, that
  allows the handling of large raw mass spectrometry experiment on
  commodity hardware and illustrate how the package is used for
  elegant data processing, method development and visualisation.
\end{abstract}

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many more. \texttt{MSnbase}
\textcolor{red}{relies on the \texttt{mzR}
  package~\cite{Chambers:2012} to import raw mass spectrometry data
  from one of the many community-maintained open standards formats
  (mzML, mzXML, mzdata or ANDI-MS/netCDF) and} provides a rich
interface to manipulate such objects. The code chunk below illustrates
such an object as displayed in the R console and an enumeration of the
metadata fields.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
either in-memory or on-disk.

@@

<<filt>>=
load("bench_t_filt.rda")
p_filt <- 
    tibble(time = microbenchmark:::convert_to_unit(t_filt$time, "ms"),
           expr = as.character(t_filt$expr)) %>%    
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = mode, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Filtering time [millisec]") + 
    xlab("Backend") +    
    theme(legend.position = "none")

@@

<<access>>=
load("bench_t_access.rda")
p_access <- 
    tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
           expr = as.character(t_access$expr)) %>%
    mutate(n = sub("^access_.+_", "", expr)) %>%
    mutate(n = sub("all", 6103, n)) %>%
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = n, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Raw data access [millisec]") +
    xlab("Number of spectra (out of 6103)") +
    facet_wrap(~ mode) +
    theme(legend.position = "none")
@

@@

\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
  direct data manipulations are not possible. We thus implemented a \textit{lazy
  processing} mechanism in this backend that caches any data manipulation
operation in a processing queue in the object itself. Only when the user
accesses m/z or intensity values these operations are applied}.
\sout{The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing} in the object itself.} Operations on the raw
data are stored in a processing queue and only effectively applied
when the user accesses m/z or intensity values. \textcolor{red}{As an additional
  advantage, operations on subsets of the data become much faster since data
  manipulations are applied only to data subsets instead of the full data set
  at once. Also, data access in on-disk backends is parallelized by data file
  ensuring a higher performance of this backend over conventional in-memory data
  representations.} 
As an example, the
following short analysis pipeline, that can equally be applied to
in-memory or on-disk data, retains MS2 spectra acquired between 1000
and 3000 seconds, extracts the m/z range corresponding to the TMT
6-plex range and focuses on the MS2 spectra with a precursor intensity
greater than $11 \times 10^6$ (the median precursor intensity).

@@

The \texttt{MSnExp} data structure and its interface constitute a
efficient prototyping environment for computational method
development. We illustrate this by demonstrating how to implement the
BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
acquisition aims at improving the detection of intact precursor ions
by distributing the charge capacity over multiple narrow m/z segments
and thus limiting the proportion of highly abundant precursors in each
segment. A full scan is reconstructed by combining the respective
adjacent segments of the BoxCar acquisitions. The
\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
that demonstrates this. The simple method is composed of three steps
is described below and illustrated with code from
\texttt{MSnbaseBoxCar} in the following code chunk.

@@

<<bc1, echo=TRUE, eval=FALSE>>=
bc <- readMSData("boxcar.mzML", mode = "onDisk") %>%
      bc_groups() %>%       ## identify BoxCar groups (bc_groups)
      filterBoxCar() %>%    ## keep only BoxCar spectra
      bc_zero_out_box() %>% ## remove signal outside of BoxCar segments
      combineSpectra(fcol = "bc_groups", ## reconstruct full spectrum
                     method = boxcarCombine)
@

@@

\begin{figure}[p]
  \centering
<<plot_boxcar, message=FALSE, warning=FALSE, fig.height=8>>=
load("boxcar.rda")
p1 + p2 + p3 + p4 + 
    plot_layout(ncol = 1, 
                heights = c(0.7, 1.2, 0.7, 0.7)) + 
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10),
          axis.text = element_text(size = 6),
          plot.margin = margin(0, 0, 0, 0, "mm"))
@
\caption{BoxCar processing with \texttt{MSnbase}. (a) Standard full
  scan with (b) three corresponding BoxCar scans showing the adjacent
  segments. Figure (c) shows the overlapping intact BoxCar segments
  and (d) the same segments after cleaning, i.e. where peaks outside
  of the segments were removed. The reconstructed full scan is shown
  on panel (e). Spectra visualisation, as shown here, rely on the
  \texttt{ggplot2} \cite{ggplot2} package. }
\label{fig:bc}
\end{figure}

@@

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
herein. According to \texttt{MSnbase}'s Bioconductor page, there are
36 Bioconductor packages that depend, import or suggest it. Among
these are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
spectrometry-based spatial proteomics data,
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
statistical analysis for quantitative proteomics data,
\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
MS files and build MassBank records,
\texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
suitability monitoring and quality control of targeted proteomic
experiments and the widely used \texttt{xcms}~\cite{Smith:2006} package
for the processing and analysis of metabolomics data. \texttt{MSnbase}
is also used in non-R/Bioconductor software, such as for example
IsoProt~\cite{Griss:2019}, that provides a reproducible workflow for
iTRAQ/TMT experiments. \texttt{MSnbase} currently ranks 101 out of
1823 packages based on the monthly downloads from unique IP addresses,
tallying over 1000 downloads from unique IP addresses each months.

>>>>

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. We focus on the new \textit{on-disk} infrastructure, that
  allows the handling of large raw mass spectrometry experiment on
  commodity hardware and illustrate how the package is used for
  elegant data processing, method development, and visualisation.
\end{abstract}

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a
\textcolor{red}{single} table with variables along the columns and
each row associated to a spectrum. Among the metadata available for
each spectrum, there are MS level, acquisition number, retention time,
precursor m/z and intensity (for MS level 2 and above), and many
more. \texttt{MSnbase} \textcolor{red}{relies on the \texttt{mzR}
  package~\cite{Chambers:2012} to import raw mass spectrometry data
  from one of the many community-maintained open standards formats
  (mzML, mzXML, mzData or ANDI-MS/netCDF) and} provides a rich
interface to manipulate such objects. The code chunk below illustrates
such an object as displayed in the R console and an enumeration of the
metadata fields.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
  data from an mzML file and} storing \sout{data}\textcolor{red}{them} 
either in-memory or on-disk.

@@

<<filt>>=
load("bench_t_filt.rda")
p_filt <- 
    tibble(time = microbenchmark:::convert_to_unit(t_filt$time, "ms"),
           expr = as.character(t_filt$expr)) %>%    
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = mode, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Filtering time [ms]") + 
    xlab("Backend") +    
    theme(legend.position = "none")

@@

<<access>>=
load("bench_t_access.rda")
p_access <- 
    tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
           expr = as.character(t_access$expr)) %>%
    mutate(n = sub("^access_.+_", "", expr)) %>%
    mutate(n = sub("all", 6103, n)) %>%
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = n, y = time, fill = mode)) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Raw data access [ms]") +
    xlab("Number of spectra (out of 6103)") +
    facet_wrap(~ mode) +
    theme(legend.position = "none")
@

@@

\textcolor{red}{Because the on-disk backend does not hold all the data
  in memory, direct data manipulations are not possible. We thus
  implemented a \textit{lazy processing} mechanism for this backend
  that caches any data manipulation operations in a processing queue
  in the object itself. Only when the user accesses m/z or intensity
  values are these operations are applied}.  \sout{The on-disk backend
  also offers efficient data manipulation by way of \textit{lazy
    processing} in the object itself.} Operations on the raw data are
stored in a processing queue and only effectively applied when the
user accesses m/z or intensity values. \textcolor{red}{As an
  additional advantage, operations on subsets of the data become much
  faster since data manipulations are applied only to data subsets
  instead of the full data set at once. Also, data access in on-disk
  backends is parallelized by data file ensuring a higher performance
  of this backend over conventional in-memory data representations.}
As an example, the following short analysis pipeline, that can equally
be applied to in-memory or on-disk data, retains MS2 spectra acquired
between 1000 and 3000 seconds, extracts the m/z range corresponding to
the TMT 6-plex range and focuses on the MS2 spectra with a precursor
intensity greater than $11 \times 10^6$ (the median precursor
intensity).

@@

The \texttt{MSnExp} data structure and its interface constitute an
efficient prototyping environment for computational method
development. We illustrate this by demonstrating how to implement the
BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
acquisition aims at improving the detection of intact precursor ions
by distributing the charge capacity over multiple narrow m/z segments
and thus limiting the proportion of highly abundant precursors in each
segment. A full scan is reconstructed by combining the respective
adjacent segments of the BoxCar acquisitions. The
\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
that demonstrates this. The simple method is composed of three steps
is described below and illustrated with code from
\texttt{MSnbaseBoxCar} in the following code chunk.

@@

<<bc1, echo=TRUE, eval=FALSE>>=
bc <- readMSData("boxcar.mzML", mode = "onDisk") %>%
      bc_groups() %>%       ## identify BoxCar groups (creates 'bc_groups')
      filterBoxCar() %>%    ## keep only BoxCar spectra
      bc_zero_out_box() %>% ## remove signal outside of BoxCar segments
      combineSpectra(fcol = "bc_groups", ## reconstruct full spectrum
                     method = boxcarCombine)
@

@@

\begin{figure}[p]
  \centering
<<plot_boxcar, message=FALSE, warning=FALSE, fig.height=8>>=
load("boxcar.rda")
p4 <- p4 + xlab("m/z")
p1 + p2 + p3 + p4 + 
    plot_layout(ncol = 1, 
                heights = c(0.7, 1.2, 0.7, 0.7)) + 
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10),
          axis.text = element_text(size = 6),
          plot.margin = margin(0, 0, 0, 0, "mm"))
@
\caption{BoxCar processing with \texttt{MSnbase}. (a) Standard full
  scan with (b) three corresponding BoxCar scans showing the adjacent
  segments. Figure (c) shows the overlapping intact BoxCar segments
  and (d) the same segments after cleaning, i.e. where peaks outside
  of the segments were removed. The reconstructed full scan is shown
  on panel (e). Spectra visualisation, as shown here, rely on the
  \texttt{ggplot2} \cite{ggplot2} package. }
\label{fig:bc}
\end{figure}

@@

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
herein. According to \texttt{MSnbase}'s Bioconductor page, there are
36 Bioconductor packages that depend, import or suggest it. Among
these are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
spectrometry-based spatial proteomics data,
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
statistical analysis \sout{for} \textcolor{red}{of} quantitative
proteomics data, \texttt{RMassBank}~\cite{Stravs:2013} to process
metabolomics tandem MS files and build MassBank records,
\texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
suitability monitoring and quality control of targeted proteomic
experiments and the widely used \texttt{xcms}~\cite{Smith:2006}
package for the processing and analysis of metabolomics
data. \texttt{MSnbase} is also used in non-R/Bioconductor software,
such as for example IsoProt~\cite{Griss:2019}, that provides a
reproducible workflow for iTRAQ/TMT experiments. \texttt{MSnbase}
currently ranks 101 out of 1823 packages based on the monthly
downloads from unique IP addresses, tallying over 1000 downloads from
unique IP addresses each months.
